Neural Connectome of the Ctenophore Statocyst

### 2025-04-07 manuscript/Manuscript.qmd
MS: edit discussion, refs

In this study, we traced all the constituent cells of the aboral organ in five-day-old larvae of *Mnemiopsis leidyi* and completed a comprehensive neural map. As a result, we identified a novel type of syncytial neuron with a previously unreported morphology. Furthermore, we found that three such syncytial neurons project to distinct quadrants, integrating with balancer cell groups to form a gravity-sensing neural circuit. This study represents the first complete neural circuit map reported in ctenophores. Additionally, our analysis of balancer ciliary movements, based on neural circuit data, revealed differences in ciliary arrest control arising from circuit variations.

<!--#  discuss that it is a complete circuit, with no neuronal connections in or out - effector output is likely by hydrodynamics, as suggested by Tamm -->

### Identification of a Previously Unreported Syncytial Neuron

We identified three syncytial neurons with multiple nuclei in the aboral organ of five-day-old *M. leidyi* cydippid larvae. Burkhurdt et al. previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023]. However, the newly discovered syncytial neurons embedded within the aboral organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons. Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces. These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.

### Functional Diversification of the Three Syncytial Neurons

The three identified neurons may be classified into two distinct functional types based on their spatial arrangement and their influence on the differential control of balancer ciliary movement between the quadrants. In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements. For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Verasztó2017]. Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells. Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.

@@

>>>>

In this study, we generated a comprehensive neural map of the aboral organ of a five-day-old *Mnemiopsis leidyi*. We described a novel type of syncytial neuron and analysed its synaptic connections to balancer-cell groups across the four quadrants of the aboral organ. Given that all neurons are intrinsic to the aboral organ and our EM volume covers the entire organ, we could recover a complete circuit map, the first for a ctenophore. Contrary to our expectations, the nerve net has no motor connections to effector organs, such as the ciliary grooves or comb rows. Our high-speed imaging suggests that the nerve net instead coordinates the activity of balancer cilia. This agrees with a hydrodynamic model of graviorientation <!-- add ref to Tamm -->, with balancer cilia directly responding to statolyth load and acting as pacemakers for the comb rows via the ciliated grooves. The nerve net could in turn tune or reset sensitivites and valence of balancer cells, in a coordinatory role, as we will discuss below.

### A New Type of Syncytial Neuron

We identified three syncytial neurons with multiple nuclei in the aboral organ of the cydippid larva. Burkhurdt et al. previously reported syncytial neurons in the sub-epithelial nerve net (SNN) of a one-day-old larva [@Burkhardt2023]. The syncytial neurons in the aboral organ did not exhibit the characteristic "pearl necklace-like structure" of the SNN neurons. Instead, their morphology was more reminiscent of fiber cells found in Placozoa [@Mayorova2021], with cell projections extending through intercellular spaces. In our volume, we also identified two SNN-like neurons with a pearl necklace-like morphology, indicating that the differences between ANN and SNN are not due to differences in stage or sample preparation. This confirms the ANN neurons as a second distinct type of syncytial neuron in ctenophores.

### Functional Diversification of the Aboral Syncytial Neurons

The three ANN neurons form two functional types based on their spatial arrangement and differential innervation of balancer cells.

@@

In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements. For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Veraszto2017]. Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells. Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.
